### 2021-06-05 paper.tex
Add lots of introductory material

\todo{add switching function}

The helical kink instability is a form of ideal MHD instability which occurs in highly twisted magnetic flux tubes such as those making up much of the solar corona~\cite{realeCoronalLoopsObservations2014} and has been well studied in the coronal context~\cite{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003c,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020a}. Given its energetic nonlinear development, it is considered a potential mechanism for heating the solar corona through the theory of nanoflares~\cite{klimchukSolvingCoronalHeating2006a,browningMechanismsSolarCoronal1991} and a key mechanism in the production of solar flares~\cite{hoodKinkInstabilitySolar1979}. Previous work by the same authors has investigated a twisted magnetic flux tube already linearly unstable to the helical kink instability, investigating particularly the effect of anisotropic viscosity on the nonlinear dynamics~\cite{quinnEffectAnisotropicViscosity2020a}. In~\cite{quinnEffectAnisotropicViscosity2020a} and most other investigations of the kink instability, a perturbation is applied to an already significantly twisted flux tube. An alternative way to excite the instability is to start with an initially straight field and apply twisting motions at the boundaries to form a twisted flux tube which eventually becomes unstable. This kind of dynamic excitation of the kink instability is useful in that it represents more closely the real development of a magnetic flux tube and associated instabilities in the solar corona. The dynamic twisting of the flux tube reveals an additional instability, the fluting instability, which has been theorised (for example, in~\cite{priestMagnetohydrodynamicsSuna}) but not yet observed in coronal simulations or observations, to our knowledge.

The fluting instability arises in magnetised plasmas where the plasma pressure gradient is directed in the same direction as the field line curvature, that is the pressure and magnetic tension forces compete. This is similar to the competition between pressure and gravitational forces which gives rise to the Rayleigh-Taylor instability (RTI). In MHD, the RTI is a typical example of an ideal interchange instability, where magnetic field lines are minimally bent and are, instead, exchanged during the evolution of the instability. The ideal fluting instability is another example of an ideal interchange instability, but confined to a cylindrical geometry, the name ``fluting instability'' referring to its likeness to a fluted column. In a twisted flux tube like a coronal loop, the magnetic curvature is always directed towards the axis so the tube may be unstable to fluting when the pressure decreases outwards from a high-pressure core. Such a pressure distribution is found in the flux tubes studied here. The appearance of the fluting instability is illustrated by, for example, the pressure contours in figure~\ref{fig:kink_field_line_plots}, where the perturbations follow the pitch of the twisted field.

In other solar contexts, interchange instabilities can be found in the form of ballooning modes in arcades~\cite{hoodBallooningInstabilitiesSolar1986}, as the instability which forms tubes of specific size in the photosphere~\cite{bunteInterchangeInstabilitySolar1993} and in the buoyancy of flux tubes~\cite{schuesslerInterchangeInstabilitySmall1984}. However, the instability is more commonly studied in fusion contexts.

While the stability of both the ideal and resistive fluting instabilities has been studied extensively in fusion research~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}, the focus is generally on understanding how a particular plasma device may be stabilised to the instability in particular geometries such as that of the mirror machine~\cite{jungwirthTheoryFluteInstability1965} or in toroidal geometries such as the tokomak~\cite{shafranovFluteInstabilityCurrentcarrying1968}. The resistive fluting instability (also known as the resistive interchange instability) can be excited even when the ideal fluting instability is stabilised. As a result, this has been given significantly more attention~\cite{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}. While this body of research is useful and applicable in solar contexts, it is mostly limited to the stability and linear development of the fluting instability. More detailed investigations of its nonlinear development is required to understand its importance in the context of coronal dynamics and coronal heating. The development of the fluting instability and its interaction with the simultaneously growing kink instability is the focus of this work and the experiments described here represent an initial exploration into the nonlinear fluting instability in the solar corona.

@@

The non-dimensionalisation scheme and values are identical to those used in~\cite{quinnEffectAnisotropicViscosity2020a}. The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of dimension $[-2,2]^3$. The velocity is set everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$, the internal energy to $\varepsilon = 8.67\times 10^{-4}$ (corresponding to a temperature of $10^6$ K). At the boundaries, the magnetic field, density, and energy are fixed to their initial values and the velocity to zero except at the upper and lower boundary where the twisting driver, described below, is prescribed. The spatial derivatives of these variables are also set to zero at the boundaries. The resolution is $512$ grid points per dimension, comparable to the highest resolution kink instability studies of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies of~\cite{barefordShockHeatingNumerical2015}. The switching function employed is the von Mises function~\eqref{eq:switching_function} and the switching parameter is chosen to be $a_0 = 150$, a value sufficiently large to ensure the viscosity is anisotropic throughout the entire domain.

@@

The flux rope is formed by prescribing a slowly accelerating, rotating flow at the upper $z$-boundary as
\begin{equation}
  \label{eq:null_twisting_profile}
  \vec{u} = u_0 u_r(r) u_t(t) (-y, x, 0)^T,
\end{equation}
where $u_r(r)$ describes the radial profile of the twisting motion in terms of the radius $r^2 = x^2 + y^2$,
\begin{equation}
  \label{eq:radial_twisting_function}
  u_r(r) = u_{r0}(1 + \tanh(1 - r_d r^2)),
\end{equation}
where $r_d$ controls the radial extent of the driver, $u_{r0}$ is a normalising factor, and $u_t(t)$ describes the imposed acceleration of the twisting motion,
\begin{equation}
  \label{eq:ramping_up_function}
  u_t(t) = \tanh^2(t/t_r),
\end{equation}
where the parameter $t_r$ controls the time taken to reach the final driver velocity $u_0$. The functions $u_r(r)$ and $u_t(t)$ are plotted in figure~\ref{fig:kink_driver}. At the lower boundary, the flow is in the opposite direction.

This form of driver allows the system to be accelerated slowly enough that the production of disruptive shocks and fast waves is minimal. It is unavoidable that some waves are produced during the boundary acceleration, however these usefully provide one source of noise which eventually forms a perturbation.

The driver velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0} = 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$, resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en times. These driver parameters correspond to a peak rotational period of $T_R = 15.92$, the length of time taken for one full turn to be injected by a single driver. Both drivers result in twist being added at a rate of $2\pi$ every $7.96$ Alfv\'en times. The twist profile across the entire flux tube develops in such a way that it is qualitatively similar to that studied in chapter~\cite{quinnEffectAnisotropicViscosity2020a} by $t\approx 20$, however the length of the flux tubes differs significantly. This configuration produces a $z$-directed tube of increasingly twisted magnetic field that eventually becomes unstable to both the fluting instability and the helical kink instability.

Two pairs of simulations were performed, one pair where the background resistivity is set to $\eta=10^{-3}$ and another where $\eta=10^{-4}$. As in chapter~\cite{quinnEffectAnisotropicViscosity2020a}, only background resistivity is used. Each simulation pair consists of one simulation using isotropic viscosity and another using the switching model~\eqref{eq:switching_model_simple} with the von Mises switching function~\eqref{eq:switching_function}. The value of viscosity is set to $\nu = 10^{-4}$ in all cases.

>>>>

The helical kink instability is a form of ideal magnetohydrodynamic (MHD) instability which occurs in highly twisted magnetic flux tubes such as those making up much of the solar corona~\cite{realeCoronalLoopsObservations2014} and has been well studied in the coronal context~\cite{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003c,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020a}. Given its energetic nonlinear development, it is considered a potential mechanism for heating the solar corona through the theory of nanoflares~\cite{klimchukSolvingCoronalHeating2006a,browningMechanismsSolarCoronal1991} and a key mechanism in the production of solar flares~\cite{hoodKinkInstabilitySolar1979}. Previous work by the same authors has investigated a twisted magnetic flux tube already linearly unstable to the helical kink instability, investigating particularly the effect of anisotropic viscosity on the nonlinear dynamics~\cite{quinnEffectAnisotropicViscosity2020a}. In~\cite{quinnEffectAnisotropicViscosity2020a} and most other investigations of the kink instability, a perturbation is applied to an already significantly twisted flux tube. An alternative way to excite the instability (and the way employed here) is to start with an initially straight field and apply twisting motions at the boundaries to form a twisted flux tube which eventually becomes unstable. This kind of dynamic excitation of the kink instability is useful in that it represents more closely the real development of a magnetic flux tube and associated instabilities in the solar corona. The dynamic twisting of the flux tube reveals an additional instability, the fluting instability, which has been theorised (for example, in~\cite{priestMagnetohydrodynamicsSuna}) but not yet observed in coronal simulations or observations, to our knowledge.

The fluting instability arises in magnetised plasmas where the plasma pressure gradient is directed in the same direction as the field line curvature, that is the pressure and magnetic tension forces compete. This is similar to the competition between pressure and gravitational forces which gives rise to the Rayleigh-Taylor instability (RTI). In MHD, the RTI is a typical example of an ideal interchange instability, where magnetic field lines are minimally bent and are, instead, exchanged during the evolution of the instability. The ideal fluting instability is another example of an ideal interchange instability but confined to a cylindrical geometry, the name ``fluting instability'' referring to its likeness to a fluted column. In a twisted flux tube like a simple, unbraided coronal loop, the magnetic curvature is always directed towards the axis so the tube may be unstable to fluting when the pressure decreases outwards from a high-pressure core. Such a pressure distribution is generated in the flux tubes studied here as a result of the driving. The appearance of the fluting instability is illustrated by, for example, the pressure contours in figure~\ref{fig:kink_field_line_plots}, where the perturbations follow the pitch of the twisted field.

In other solar contexts, interchange instabilities can be found in the form of ballooning modes in arcades~\cite{hoodBallooningInstabilitiesSolar1986}, as the instability which forms tubes of specific size in the photosphere~\cite{bunteInterchangeInstabilitySolar1993} and in the buoyancy of flux tubes~\cite{schuesslerInterchangeInstabilitySmall1984}. However, the fluting instability specifically is more commonly studied in fusion contexts~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}. In fusion, the focus is generally on understanding how a particular plasma device may be stabilised to the instability in particular geometries such as that of the mirror machine~\cite{jungwirthTheoryFluteInstability1965} or in toroidal geometries such as the tokomak~\cite{shafranovFluteInstabilityCurrentcarrying1968}. The resistive fluting instability (also known as the resistive interchange instability) can be excited even when the ideal fluting instability is stabilised. As a result, this has been given significantly more attention~\cite{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}. While this body of research is useful and applicable in solar contexts, it is mostly limited to the study of the stability and linear development of the fluting instability, the nonlinear development being of secondary importance to the functioning of fusion devices. More detailed investigations of its nonlinear development is required to understand its importance in the context of coronal dynamics and coronal heating. The development of the fluting instability and its interaction with the simultaneously growing kink instability is the focus of this work and the experiments described here represent an initial exploration into the nonlinear fluting instability in the solar corona. Of particular interest is the effect of anisotropic viscosity, which is found to strongly affect the growth of the fluting instability.

It is well known that viscosity in magnetised plasmas (such as that which makes up the solar corona) is anisotropic and strongly dependent on the strength and direction of the local magnetic field~\cite{hollwegViscosityChewGoldbergerLowEquations1986a,hollwegViscosityMagnetizedPlasma1985,Braginskii1965}. The development of the switching model~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017} and implementation in Lare3d has allowed the investigation of the effect of anisotropic viscosity on solar phenomena. We find the stability of Lare3d without (the more commonly employed) isotropic or shock viscosity to be sufficient for at least the study of kink instabilities in coronal loops~\cite{quinnEffectAnisotropicViscosity2020a}. The interest in anisotropic viscosity stems from the unanswered question of which heating mechanism (viscous or Ohmic) is dominant in the solar corona~\cite{klimchukSolvingCoronalHeating2006a}, an important facet of solving the coronal heating problem. Using scaling laws, it has been suggested that viscous heating (generated through anisotropic viscosity) can dwarf that of Ohmic heating~\cite{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}. However, due to computational and observational limitations, this cannot be directly tested, and so the influence of other factors such as small scale instabilities and turbulence is relatively unknown~\cite{klimchukSolvingCoronalHeating2006a}. In addition to directly heating the plasma, viscosity plays a part in the damping of instabilities and waves~\cite{rudermanSlowSurfaceWave2000a}. It is this effect we are most interested in here, and it shall be found that the use of anisotropic viscosity permits the growth of the fluting instability, which is otherwise strongly damped by isotropic viscosity.

@@

Lare3d numerically solves the MHD equations for the density $\rho$, plasma velocity $\vec{u}$, pressure $p$, magnetic field $\vec{B}$ and internal energy $\varepsilon$,
\begin{gather}
%\begin{equation}
\label{eq:mhda}
\frac{D\rho}{Dt} = - \rho \vec{\nabla} \cdot \vec{u},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\vec{u}}{Dt} = -\vec{\nabla} p + \vec{\jmath} \times \vec{B} + \vec{\nabla} \cdot \ten{\sigma},\\
%\end{equation}
%\begin{equation}
\frac{D\vec{B}}{Dt} = (\vec{B} \cdot \vec{\nabla})\vec{u} - (\vec{\nabla} \cdot \vec{u})\vec{B} + \eta \nabla^2 \vec{B},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\varepsilon}{Dt} = -p \vec{\nabla} \cdot \vec{u} + {Q}_{\nu} + {Q}_{\eta},%\\
\label{eq:energy}
%\end{equation}
\end{gather}
where $\eta$ is the non-dimensionalised resistivity, $\jmath = \nabla \times \vec{B}$ is the current density and $\ten{\sigma}$ is the viscous stress tensor (described below). The terms ${Q}_{\nu} = \ten{\sigma} : \vec{\nabla}\vec{u}$ and ${Q}_{\eta} = \eta | \vec{\jmath} |^2$ are viscous heating and Ohmic heating, respectively. The system is closed by the inclusion of the equation of state for an ideal gas
\begin{equation}
\varepsilon = \frac{p}{\rho(\gamma - 1)},
\end{equation}
where the specific heat ratio is given by $\gamma = 5/3$.

The isotropic (or Newtonian) viscosity stress tensor is given by
\begin{equation}
  \label{eq:isotropic_viscous_tensor}
  \ten{\sigma} = \nu \ten{W},
\end{equation}
where $\nu$ is the viscous transport parameter, generally referred to as the viscosity,
\begin{equation}
  \label{eq:rate_of_strain}
  \ten{W} = \nabla\vec{u} + (\nabla\vec{u})^T - \tfrac{2}{3}(\nabla \cdot \vec{u})\ten{I},
\end{equation}
is the rate of strain tensor, and $\ten{I}$ is the  $3\times 3$ identity matrix. We use an identical model of anisotropic viscosity as that used in previous work~\cite{quinnEffectAnisotropicViscosity2020a}, the switching model, first presented in~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}. While this model is intended to revert to an isotropic model in the vicinity of magnetic null points, we effectively only make use of the anisotropic component, since there are no null points present in the simulations at any point. To this end, we set the switching parameter (which controls the degree of viscous anisotropy) to $a_0 = 150$ in Lare3d's control, a value sufficiently large that the anisotropic viscous stress tensor is given by
\begin{equation}
  \label{eq:switching_model}
\ten{\sigma}_{\text{swi}} = \nu \left[\frac{3}{2}(\ten{W}\vec{b}\cdot\vec{b}) \left( \vec{b} \otimes \vec{b} - \frac{1}{3}\ten{I} \right)\right],
\end{equation}
where $\vec{b}$ is the unit vector in the direction of the magnetic field. This tensor is identical to the strong field approximation of Braginskii~\cite{braginskiiTransportProcessesPlasma1965}. Full use of the switching model can be found in~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020a}\todo{ref KHI paper}.

The non-dimensionalisation scheme and values align with the scheme used by Lare3d~\cite{arberStaggeredGridLagrangian2001} and are identical to those used in~\cite{quinnEffectAnisotropicViscosity2020a}. The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of dimension $[-2,2]^3$. The velocity is set everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$, the internal energy to $\varepsilon = 8.67\times 10^{-4}$ (corresponding to a temperature of $10^6$ K). \todo{what's the plasma beta?} At the boundaries, the magnetic field, density, and energy are fixed to their initial values and the velocity to zero except at the boundaries where the twisting driver, described below, is prescribed. The spatial derivatives of these variables are also set to zero at the boundaries. The resolution is $512$ grid points per dimension, comparable to the highest resolution kink instability studies of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies of~\cite{barefordShockHeatingNumerical2015}.

@@

The flux rope is formed by prescribing a slowly accelerating, rotating flow at the upper $z$-boundary as
\begin{equation}
  \label{eq:null_twisting_profile}
  \vec{u} = u_0 u_r(r) u_t(t) (-y, x, 0)^T,
\end{equation}
where $u_r(r)$ describes the radial profile of the twisting motion in terms of the radius $r^2 = x^2 + y^2$,
\begin{equation}
  \label{eq:radial_twisting_function}
  u_r(r) = u_{r0}(1 + \tanh(1 - r_d r^2)),
\end{equation}
where $r_d$ controls the radial extent of the driver, $u_{r0}$ is a normalising factor, and $u_t(t)$ describes the imposed acceleration of the twisting motion,
\begin{equation}
  \label{eq:ramping_up_function}
  u_t(t) = \tanh^2(t/t_r),
\end{equation}
where the parameter $t_r$ controls the time taken to reach the final driver velocity $u_0$. The functions $u_r(r)$ and $u_t(t)$ are plotted in figure~\ref{fig:kink_driver}. At the lower boundary, the flow is in the opposite direction. This form of driver allows the system to be accelerated slowly enough that the production of disruptive shocks and fast waves is minimal. It is unavoidable that some waves are produced during the boundary acceleration, however these usefully provide one source of noise which eventually forms a perturbation.

The driver velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0} = 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$, resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en times. These driver parameters correspond to a peak rotational period of $T_R = 15.92$, the length of time taken for one full turn to be injected by a single driver. Both drivers result in twist being added at a rate of $2\pi$ every $7.96$ Alfv\'en times. The twist profile across the entire flux tube develops in such a way that by $t\approx 20$, it is qualitatively similar to those studied in~\cite{quinnEffectAnisotropicViscosity2020a,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015} however the length of the flux tubes differs significantly. This configuration produces a $z$-directed tube of increasingly twisted magnetic field that eventually becomes unstable to both the fluting instability and the helical kink instability.

Two pairs of simulations were performed, one pair where the background resistivity is set to $\eta=10^{-3}$ and another where $\eta=10^{-4}$. As in~\cite{quinnEffectAnisotropicViscosity2020a}, only background resistivity is used. Each simulation pair consists of one simulation using isotropic viscosity~\eqref{eq:isotropic_viscous_tensor} and another using the switching model~\eqref{eq:switching_model}. The value of viscosity is set to $\nu = 10^{-4}$ in all cases.
